Mixture Models for Single--Cell Assays with Application to Vaccine Studies

### 2012-09-11 MIMOSA.tex
Reformat for Biometrics

%\documentclass[11pt]{article}
\documentclass[useAMS,referee,usenatbib]{biom}
\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}
%\usepackage{geometry,authblk,multicol,amsmath}
\usepackage{amsmath,tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

%\usepackage[square,numbers,sort&compress]{natbib}
%\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{url}

@@

\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
%\author[1]{Greg Finak}
%\author[2]{Steve De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}

%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, Steve De Rosa$^{1,2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

%\date{\today}

\begin{document}
%\maketitle
\date{{\it Received August} 2012.}

@@

\label{firstpage}

\begin{abstract}
Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
\end{keywords}

\maketitle

\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers). 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination). Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (e.g. animal studies).

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders} whereas subject that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects. 
This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows, Section \ref{s:data} introduces the data and notations used in the paper. In Section \ref{s:DEone}, we present our model for testing differential marker expression in the univariate case. Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model. In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using a single-cell gene expression data. Finally, in Section \ref{s:discussion} we discuss our findings and future work.

\section{Data structure and notation} 
\label{s:data}
In this paper we consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen with 120 subjects \citep{Goepfert:2011ci}. 
The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points, day 0 and day 182. 
Three cytokines (IFN$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

In the remainder of the paper, we use the following notation to describe our model.  From this point on, we assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a marker. Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample. For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

\section{Differential expression with one marker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. Here, by univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$). Without loss of generality, we treat the univariate case as a one marker case (\textit{i.e.} $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

\begin{table}
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Marker}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}

For a given subject and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects, we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.

>>>>

%\documentclass[11pt]{article}
\documentclass[useAMS,referee,usenatbib]{biom}
\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
%\usepackage{natbib}
\newcommand{\tr}{\mbox{tr}}
%\usepackage[margin=1in]{geometry}
%\usepackage{authblk}
%\usepackage{multicol}
%\usepackage{tikz}
\usepackage{amsmath,tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

%\usepackage[round,comma]{natbib}
%\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{booktabs}
%\usepackage{multirow}
\usepackage{url}

@@

%\title[MIMOSA: Mixture Models for Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
\title[MIMOSA: Mixture Models for Single--Cell Assays]{Mixture Models for Single--Cell Assays with Application to Vaccine Studies}

%\author[1]{Greg Finak}
%\author[1]{Andrew McDavid}
%\author[3]{Pratip Chattopadhyay}
%\author[3]{Maria Dominguez}
%\author[1,2]{Steve De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}
%
%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

\author{Greg Finak$^{1}$, \and
Andrew McDavid$^{1}$,
Pratip Chattopadhyay$^{3}$,
Maria Dominguez$^{3}$, 
Steve De Rosa$^{1,2}$,
Mario Roederer$^{3}$,
\and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

%\date{\today}                                       
\begin{document}

\date{{\it Received XXXXX} 2012.}

@@

\label{firstpage}
%\maketitle
\begin{abstract}
Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against frequentist approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
\end{keywords}
\maketitle

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers). 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. proteins, genes, cytokines, etc.) per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level biomarker fluorescence intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers. 
Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination). Note that we use the term `subject' throughout the paper, but the approaches described are general and can be applied to other experimental units (\textit{e.g.}, animal studies).

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. The antigen--specific T-cell subpopulations (i.e. the subset that can respond to one specific antigen) constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic biomarkers (CD3, CD4, and CD8), used to subset T-cells, and functional biomarkers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing ICS assays, current methods range from ad-hoc rules based on log-fold changes~\citep{Trigona:2003}, to non--parametric methods~\citep{Sinclair:2004hs}, to permutation tests based on Hotelling's T$^2$ statistics~\citep{Nason:2006dx}, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Horton:2007tsa,Proschan:2009ks,Peiperl:2010ej,Nason:2006dx}.
All of these methods test subjects separately, and no information is shared across observations even though one could expect some similarities across responders (or non-responders).

The framework developed in this paper, named MIMOSA (Mixture Models for Single Cell Assays), addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects. 
This sharing of information helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increases sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be modelled simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows; Section \ref{s:data} introduces the data and notations used in the paper. In Section \ref{s:DEone}, we present our model for testing differential biomarker expression in the univariate case. Section \ref{s:results} compares our approach to alternative methods and tests the robustness of our model. In Section \ref{s:demarkercombos} we present a multivariate extension of our model that can be used to test multivariate biomarker differential expression and present some results using a single-cell gene expression data. Finally, in Section \ref{s:discussion} we discuss our findings and future work.

\section{Notation and Data} 
\label{s:data}

In the remainder of the paper, we use the following notation to describe our model.  We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a biomarker. Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1,\dots, 2^K, i=1,\dots,I$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample, respectively. For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

We consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}. 
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points: day 0 and day 182. 
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (here we focus on HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls.

Although the immunological experiments described above will often look at multiple antigens or stimulations, in the models presented here we consider only one stimulation (i.e. antigen or condition) at a time vs. unstimulated. The issue of multiple antigens is handled through multiple testing correction.

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. Here, by univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ and/or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

\begin{table}
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for biomarker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Biomarker}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model as described in what follows.

\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows:

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix B, and we refer to it as the one-sided model.
